---
title: "Building a trajectory for use with PPR"
author: "Moi Taiga Nicholas"
date: "`r Sys.Date()`"
output: github_document
---

<!-- trajectoryTutorial.md is generated from trajectoryTutorial.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/TRAJ-",
  out.width = "100%",
  echo = TRUE,
  eval = FALSE
)
```

# Building a trajectory for use with PPR
## Introduction
This tutorial covers how to build a trajectory for use with PPR. 
We will use cellxgene to download single-cell RNA-seq data, 
We will use `Seurat` to perform basic preprocessing and dimensionality reduction.
We will use the `slingshot` package to infer the trajectory. 
We will use `GeneSwitches` to identify switching genes.

## activate the shared conda environment
```bash
conda activate /mainfs/ddnb/hack_days/mar25/conda/pprTraj
```

## Load the Required Packages
```{r install_packages, eval=FALSE}
library("Seurat")
library("SingleCellExperiment")
library("slingshot")
library("GeneSwitches")
#library("PathPinpointR")
library(devtools)
load_all()
```

## Downloading Data from CellxGene
Use `cellxgene` to retrieve data manually,
Choose a dataset which represents a simple trajectory,
download the data in the form of a Seurat object.

## load data
```{r load_data}
# load the data
seu <- readRDS("/mainfs/ddnb/hack_days/mar25/data/microgliaCSF1R.rds")


# # set the identity class to Braak.stage
# Idents(seu) <- "Braak.stage"
# # downsample the data
# seu <- subset(x = seu, downsample = 100)
```

## Basic Seurat Preprocessing
Run standard Seurat preprocessing steps: (if neccecary)

```{r preprocessing}
# view the umap using seurat
DimPlot(seu, reduction = "umap", group.by = "Braak.stage")

# seu <- NormalizeData(seu)
# seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 2000)
# seu <- ScaleData(seu)
# seu <- RunPCA(seu)
```

## Convert Seurat object to SingleCellExperiment
```{r convert_to_sce}
sce    <- SingleCellExperiment(assays = list(expdata = seu@assays$RNA$counts))
colData(sce) <- DataFrame(seu@meta.data)
reducedDims(sce)$UMAP <- seu@reductions$umap@cell.embeddings
```


## Trajectory Inference with Slingshot

```{r slingshot}
sce  <- slingshot(sce,
                  clusterLabels = "Braak.stage",
                  #start.clus  = "2",
                  #end.clus = "1",
                  reducedDim = "UMAP")

#Rename the Pseudotime column to work with GeneSwitches
colData(sce)$Pseudotime <- sce$slingPseudotime_1

# plot the trajectory
library("RColorBrewer")
# Generate colors
colors <- colorRampPalette(brewer.pal(11, "Spectral")[-6])(100)
plotcol <- colors[cut(sce$slingPseudotime_1, breaks = 100)]
# Plot the data
plot(reducedDims(sce)$UMAP, col = plotcol, pch = 16, asp = 1)
lines(SlingshotDataSet(sce), lwd = 2, col = "black")
```


## Identify switching genes with GeneSwitches
```{r geneSwitches}
# Identify a cutoff for binarization, by plotting a histogram of gene expression
hist(as.numeric(as.matrix(assays(sce)$expdata)),
     freq = FALSE,
     breaks = 1000,
     main = "Histogram of expression values",
     xlab = "Expression value",
     ylab = "Density",
     xlim = c(0, 6),
     ylim = c(0, 1))
# vertical line at 1
abline(v = 0.5, col = "red")

# using the cutoff of 0.5, binarize the expression data
sce <- binarize_exp(sce,
                    fix_cutoff = TRUE,
                    binarize_cutoff = 0.5,
                    ncores = 1)

# Find the switching point of each gene in the sce
sce <- find_switch_logistic_fastglm(sce,
                                    downsample = FALSE,
                                    show_warning = FALSE)


# visualise the switching genes
switching_genes <- filter_switchgenes(sce,
                                      allgenes = TRUE,
                                      r2cutoff = 0)

# Plot the timeline using plot_timeline_ggplot
plot_timeline_ggplot(switching_genes,
                     timedata = colData(sce)$Pseudotime,
                     txtsize = 3)

# save as rds 
saveRDS(sce, "/mainfs/ddnb/hack_days/mar25/data/TMPsce.rds")
saveRDS(switching_genes, "/mainfs/ddnb/hack_days/mar25/data/TMPswitching_genes.rds")

# load from rds
sce <- readRDS("/mainfs/ddnb/hack_days/mar25/data/TMPsce.rds")
switching_genes <- readRDS("/mainfs/ddnb/hack_days/mar25/data/TMPswitching_genes.rds")

```

## Filter switching genes further 
```{r filter_switching_genes}
#run PPR:precison
precision <- precision(sce, plot = FALSE)
# zoom in
precision <- precision(sce, n_sg_range = seq(50, 170, 10), plot = FALSE)
#
precision <- precision(sce, n_sg_range = seq(80, 120, 2), plot = FALSE)

# visualise the switching genes
switching_genes <- filter_switchgenes(sce,
                                      allgenes = TRUE,
                                      r2cutoff = 92)
```

## Save the trajectory
```{r save_trajectory}
# the information we need for PPR can be contained in a single table, save as csv
write.csv(switching_genes, "/mainfs/ddnb/hack_days/mar25/switching_genes.csv")
```