---
title: "blastocyst"
author: "Moi T. Nicholas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blastocyst}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)
```

This vignette will take you through running PPR. \
The data used here is an Intregrated dataset of [blastocyst data](http://). \
\
\

# Data Pre-procesing
#### Load neccecary packages
```{r}
library(Seurat)
library(ggplot2)
library(SingleCellExperiment)
library(slingshot)
library(RColorBrewer)
library(GeneSwitches)
library(devtools)
devtools::load_all()
```

## SEURAT
### Here we use this Reprogramming dataset as the reference 
```{r}
seu <- readRDS("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/blastocyst.rds")
#seu <- readRDS("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/additional_blastocyst/full_atlas.rds")
```

#### View the reference UMAP
```{r}
# p1 <- DimPlot(object = seu, 
#              reduction = "umap",
#              group.by = "orig.ident",
#              label = TRUE) +
#      ggtitle("Reference")
# ggsave(filename = "/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/dimplot_orig_ident.png", plot = p1)

# Plot and save the second DimPlot
# p2 <- DimPlot(object = seu,
#              reduction = "umap",
#              group.by = "seurat_clusters",
#              label = TRUE) +
#      ggtitle("Reference")
#ggsave(filename = "/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/dimplot_seurat_clusters.png", plot = p2)
```

### QC
```{r}
#seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "MT-")
# Visualize QC metrics as a violin plot
#VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA", "pctMT"), ncol = 3)
```

### QC
```{r}
#consider using double finder (has it aleady been run?)
#seu <- subset(seu, subset = nFeature_RNA > 2500 & pctMT < 10 & nCount_RNA < 1e+07)
#seu <- subset(seu, subset = pctMT < 11 & nCount_RNA < 1.8e+07)
# Visualize QC metrics as a violin plot
#VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA", "pctMT"), ncol = 3)
```

### We use subsets of the Reprogramming dataset as queries.
```{r}
# Split the reference data into samples

# Make a list of the unique sample names
sample_names <- unique(seu@meta.data$orig.ident)

# Make an empty list to store the samples
samples_seu <- list()

# Iterate through each sample name, make a list of subsets.
for (sample in sample_names){
  sample_seu <- subset(x = seu, subset = orig.ident %in% sample)
  samples_seu[[sample]] <- sample_seu
}
```


### Label transfer & Re-integration
####### *When doing this analysis with your own data you would more than likely need to label transfer with your referenece datatset.*
####### *As we are using subsets of our reference data as "samples" this wont be necessary.*

## SingeCellExperiment
### Convert objects to SingleCellExperiment objects
```{r}
sce    <- SingleCellExperiment(assays = list(expdata = seu@assays$RNA$counts))
colData(sce) <- DataFrame(seu@meta.data)
reducedDims(sce)$UMAP <- seu@reductions$umap@cell.embeddings

for (sample in sample_names){
  sample.sce <- SingleCellExperiment(assays = list(expdata = get(paste0(sample, ".seu"))@assays$RNA$counts))
  assign(paste0(sample, ".sce"), sample.sce)
}
```


## Slingshot

### Run slingshot on the reference data to produce a reprogramming trajectory.
```{r}
sce  <- slingshot(sce,
                  clusterLabels = "seurat_clusters",
                  start.clus  = "2",
                  end.clus = "1",
                  reducedDim = "UMAP")

#Rename the Pseudotime column to work with GeneSwitches
colData(sce)$Pseudotime <- sce$slingPseudotime_1
```

### Plot the slingshot trajectory.
```{r}
# Generate colors
colors <- colorRampPalette(brewer.pal(11, 'Spectral')[-6])(100)
plotcol <- colors[cut(sce$slingPseudotime_1, breaks = 100)]

# Open a PDF device to save the plot
pdf("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/slingshot.pdf")

# Plot the data
plot(reducedDims(sce)$UMAP, col = plotcol, pch = 16, asp = 1)
lines(SlingshotDataSet(sce), lwd = 2, col = 'black')

# Close the PDF device
dev.off()
```


## GeneSwitches
###  Choose a Binerization cutoff
###### **this would be good to automate, 
###### or learn how to use the other method of binarizing
```{r}
# hist(as.matrix(assays(sce)$expdata),
#            breaks = 1e6)
#  hist(as.matrix(seu@assays$RNA$counts), breaks = 1e6)
#  
#  hist(as.matrix(assays(sce)$expdata),
#            breaks = 1e6,
#            xlim = c(0, 50))
#  
#  hist(as.matrix(seu@assays$RNA$counts), breaks = 1e6, xlim =c(0, 50))
#  
#  trimmed_expdata <- assays(sce)$expdata[assays(sce)$expdata < 10]
#  hist(as.matrix(trimmed_expdata))
#  abline(v=1, col = "blue")
#  
#  trimmed_expdata <- seu@assays$RNA$counts[seu@assays$RNA$counts < 10]
#  hist(as.matrix(trimmed_expdata))
#  abline(v=1, col = "blue")
#cutoff =1
 
```

### Binarize
```{r}
#sce    <- binarize_exp(sce, fix_cutoff = TRUE, binarize_cutoff = 1, ncores = 64)
#Mole21a.sce   <- binarize_exp(Mole21a.sce, fix_cutoff = TRUE, binarize_cutoff = 1, ncores = 64)
#sce <- binarize_exp(sce, ncores = 64)
#Mole21a.sce   <- binarize_exp(Mole21a.sce, ncores = 64)
```

### CHECKPOINT
##### **with how long binerizing takes you may want to choose to save objects for future use.
```{r}
 #save the reference and sample objects
 #saveRDS(sce   , "/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/reference.rds")
 #for (sample in sample_names){
 #       saveRDS(get(paste0(sample, ".sce")), paste0("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/binarized_", sample, "_sce.rds"))
 #}

# read the reference and sample objects
#sce <- readRDS("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/reference.rds")
#sample_Mole21a_binarized <- readRDS("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/binarized_Mole21a_sce.rds")

samples_binarized <- list()
for (sample in sample_names){
  samples_binarized[[sample]] <- readRDS(paste0("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/binarized_", sample, "_sce.rds"))
}

```

### fit logistic regression and find the switching pseudo-time point for each gene
```{r}
#sce <- find_switch_logistic_fastglm(sce, downsample = FALSE, show_warning = FALSE)
```

#### **This is another time consuming process, may want to choose to save objects for future use.**
```{r}
## saveRDS(sce   , "~/R Packages/mini_data/Binerized/reference_glm.rds")
sce    <- readRDS("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/switches_gastglm_blastocyst_reference_sce.rds")
```

### Filter to only include Switching Genes
```{r}
switching_genes <- filter_switchgenes(sce, allgenes = TRUE,r2cutoff = 0.26)
```

##### View all of the switching genes
```{r}
# Open a PDF device to save the plot
pdf("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/timeline_plot.pdf")

# Plot the timeline using plot_timeline_ggplot
plot_timeline_ggplot(switching_genes, timedata = colData(sce)$Pseudotime, txtsize = 3)

# Close the PDF device
dev.off()
```


# Using PPR

##### View the selected switching genes
```{r}
ppr_timeline_plot(switching_genes)
```

### Reduce the binary counts matricies of the query data to only include the selection os switching genes.
```{r}
# filter the binary counts matricies to only include the switching genes
reference_reduced <- subset_switching_genes(sce, switching_genes)
#Mole21a_reduced   <- subset_switching_genes(sample_Mole21a_binarized, switching_genes)

# Define a list to store the results
samples_reduced <- list()

# Iterate through each Seurat object in the list
for (sample_name in names(samples_binarized)){
  sample_binarized <- samples_binarized[[sample_name]]
  # susbet each sample to only include the switching genes
  sample_reduced <- subset_switching_genes(sample_binarized, switching_genes)
   # Store the result in the new list
  samples_reduced[[sample_name]] <- sample_reduced
}

```

### Produce an estimate for the position on trajectory of each gene in each cell of a sample.
```{r}
reference_ppr    <- ppr_predict_position(reference_reduced, switching_genes)
#Mole21a_ppr      <- ppr_predict_position(Mole21a_reduced, switching_genes)
# Define a list to store the ppr objects of the samples
samples_ppr <- list()

# Iterate through each Seurat object in the predicting their positons,
# on the reference trajectory, using PathPinpointR
for (sample_name in names(samples_reduced)){
  sample_reduced <- samples_reduced[[sample_name]]
  # predict the position of each gene in each cell of the sample
  sample_ppr <- ppr_predict_position(sample_reduced, switching_genes)
  # Store the result in the new list
  samples_ppr[[sample_name]] <- sample_ppr
}
```

### Accuracy
#### *As our samples are subsets of the reference dataset we can calculate the accuracy of GSS*
```{r}
# Open a PDF device to save the plot
pdf("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/reference_accuracy.pdf")
#ppr_accuracy_test(Mole21a_ppr, sce, plot = TRUE)
ppr_accuracy_test(reference_ppr, sce, plot = TRUE)
# Close the PDF device
dev.off()

for (sample_name in names(samples_ppr)){
  sample_ppr <- samples_ppr[[sample_name]]
  pdf(paste0("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/", sample_name,"_accuracy.pdf"))
  ppr_accuracy_test(sample, sce, plot = TRUE)
  # Close the PDF device
  dev.off()
}
```

## Plotting

### Plot the predicted position of each sample:
#### *Optional: include the switching point of some genes of interest:*
```{r}

# Open a PDF device to save the plot
pdf("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/reference_ppr_output.pdf")
ppr_output_plot(reference_ppr, col = "red", overlay=FALSE, label = "Reference")
# Close the PDF device
dev.off()

for (sample_name in names(samples_ppr)){
  sample_ppr <- samples_ppr[[sample_name]]
  pdf(paste0("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/", sample_name,"_ppr_output.pdf"))
  ppr_output_plot(sample_ppr, col = "red", overlay=FALSE, label = sample_name)
  # Close the PDF device
  dev.off()
}

pdf("/mainfs/ddnb/PathPinpointR/package/PathPinpointR/data/plots/samples_ppr_output.pdf")
ppr_output_plot(samples_ppr[[1]], col = "red", overlay=FALSE, label = names(samples_ppr)[1])
ppr_output_plot(samples_ppr[[2]], col = "blue", overlay=TRUE, label = names(samples_ppr)[2])
ppr_output_plot(samples_ppr[[4]], col = "purple", overlay=TRUE, label = names(samples_ppr)[4])
ppr_output_plot(samples_ppr[[5]], col = "#000000", overlay=TRUE, label = names(samples_ppr)[5])
ppr_output_plot(samples_ppr[[8]], col = "#4dd80c", overlay=TRUE, label = names(samples_ppr)[8])
# Close the PDF device
dev.off()

```




#### Plot an indervidual cell.
```{r}
ppr_cell_plot(Mole21a_ppr, cell_idx = 1)
```

```{r}
ppr_timeline_plot(switching_genes,
                  genomic_expression_traces = TRUE,
                  reduced_binary_counts_matrix = Mole21a_reduced,
                  cell_id = 1)
```

### Precision
##### Estimate the precision of PPR by running accuracy_test at a range of values
###### BUG! 
```{r}
ppr_precision(sce)
```

### Investigate the predicted position of indervidual cells:
```{r}
#
plot(Mole21a_accuracy$true_position_of_cells_timeIDX, Mole21a_accuracy$inaccuracy,
     xlab = "True Position", ylab = "Inaccuracy", main = "Inaccuracy by True Positions")

#
boxplot(Mole21a_accuracy$inaccuracy ~ Mole21a_accuracy$true_position_of_cells_timeIDX,
        xlab = "True Position", ylab = "Inaccuracy", main = "Boxplots of Inaccuracy by True Positions")
true_position_bin <- cut(Mole21a_accuracy$true_position_of_cells_timeIDX,
                             breaks = c(0,5,10,15,20,25,30,35,40,45,50,55,60,65,70,75,80.85,90,95,100))
boxplot(Mole21a_accuracy$inaccuracy ~ true_position_bin,
        xlab = "True Position", ylab = "Inaccuracy", main = "Boxplots of Inaccuracy by True Positions")

#
plot(Mole21a_accuracy$true_position_of_cells_timeIDX,Mole21a_accuracy$predicted_position_of_cells_timeIDX,
     xlab = "True Position", ylab = "Predicted Position", main = "Predicted Positions by True Positions")
segments(0,0,100,100, lwd=2, col ="green")
segments(0,10,100,110, lwd=2, col ="blue")
segments(0,-10,100,90, lwd=2, col ="blue")
segments(0,40,100,140, lwd=2, col ="red")
segments(0,-40,100,60, lwd=2, col ="red")

#
boxplot(Mole21a_accuracy$predicted_position_of_cells_timeIDX ~ Mole21a_accuracy$true_position_of_cells_timeIDX,
        xlab = "True Position", ylab = "Predicted Position", main = "Boxplots of Predicted Positions by True Positions")
segments(0,0,100,100, lwd=2, col ="green")
segments(0,10,100,110, lwd=2, col ="blue")
segments(0,-10,100,90, lwd=2, col ="blue")
segments(0,40,100,140, lwd=2, col ="red")
segments(0,-40,100,60, lwd=2, col ="red")
```


