---
title: "Reprogramming Example"
author: "Moi T. Nicholas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reprogramming_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will take you through running GSS. \
The data used here is from a Reprogramming dataset, which can be found [Here](http://hrpi.ddnetbio.com/). \
\
\


# Data Pre-procesing

#### Load neccecary packages
```{r eval=FALSE}
library(Seurat)
library(ggplot2)
library(slingshot)
library(RColorBrewer)
library(GeneSwitches)
library(GeneSwitchesScorer)
library(SingleCellExperiment)
```

## SEURAT

### Here we use this Reprogramming dataset as the reference 
```{r eval=FALSE}
big_reference.seu <- readRDS("~/R Packages/mini_data/readySeu_hrpi.rds")
```
#### Downsample if neccecary
```{r eval=FALSE}
reference.seu <- subset(x = big_reference.seu, downsample = 100)
```

#### View the reference UMAP
```{r eval=FALSE}
DimPlot(object = reference.seu, reduction = "umap", group.by = "identity" ,label = T) +
  ggtitle("Reference")
```

### Subset the reference to be a single biologically relevent trajectory
#### In this case the trajectory of fibroblasts being reprogrammed to primed cell.
```{r eval=FALSE}
reference.seu <- subset(x = reference.seu, subset = identity %in% grep("^n", reference.seu@meta.data$identity, value = TRUE), invert = TRUE)
```

#### View the reduced reference UMAP
```{r eval=FALSE}
DimPlot(object = reference.seu, reduction = "umap", group.by = "identity" ,label = T) +
  ggtitle("Reference Subset")
```

### We use subsets of the Reprogramming dataset as queries.
```{r eval=FALSE}
fibroblast.seu <- subset(x = reference.seu, subset = identity %in% "fibroblast")
mixed.seu <- subset(x = reference.seu, subset = identity %in% "mixed")
early_primed.seu <- subset(x = reference.seu, subset = identity %in% "early-primed")
primed.seu <- subset(x = reference.seu, subset = identity %in% "primed")
```


### Label transfer & Re-integration
####### *When doing this analysis with your own data you would more than likely need to label transfer and re-integrate with your referenece datatset.*
####### *As we are using subsets of our reference data as "samples" this wont be necessary.*


## SingeCellExperiment


### Convert objects to SingleCellExperiment objects
```{r eval=FALSE}
reference.sce    <- as.SingleCellExperiment(reference.seu)
fibroblast.sce   <- as.SingleCellExperiment(fibroblast.seu)
mixed.sce        <- as.SingleCellExperiment(mixed.seu)
early_primed.sce <- as.SingleCellExperiment(early_primed.seu)
primed.sce       <- as.SingleCellExperiment(primed.seu)
```


## Slingshot

### Run slingshot on the reference data to produce a reprogramming trajectory.
```{r eval=FALSE}
reference.sce  <- slingshot(reference.sce,
                            clusterLabels = "identity",
                            start.clus  = "fibroblast",
                            end.clus = "primed",
                            reducedDim = "UMAP")
```

### Plot the slingshot trajectory.
```{r eval=FALSE}
colors <- colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)
plotcol <- colors[cut(reference.sce$slingPseudotime_1, breaks=100)]
plot(reducedDims(reference.sce)$UMAP, col = plotcol, pch=16, asp = 1)
lines(SlingshotDataSet(reference.sce), lwd=2, col='black')
```


## GeneSwitches

### Set up the SingleCellExperiment objects for GeneSwitches to use. 
```{r eval=FALSE}
reference.gs    <- SingleCellExperiment(assays = List(expdata = logcounts(reference.sce)))
  colData(reference.gs)$Pseudotime <- reference.sce$slingPseudotime_1
  reducedDims(reference.gs)        <- SimpleList(UMAP = reducedDim(reference.sce, "UMAP", withDimnames=TRUE))

fibroblast.gs   <- SingleCellExperiment(assays = List(expdata = logcounts(fibroblast.sce)))
mixed.gs        <- SingleCellExperiment(assays = List(expdata = logcounts(mixed.sce)))
early_primed.gs <- SingleCellExperiment(assays = List(expdata = logcounts(early_primed.sce)))
primed.gs       <- SingleCellExperiment(assays = List(expdata = logcounts(primed.sce)))
```

###  Choose a Binerization cutoff
###### **this would be good to automate, 
###### or learn how to use the other method of binarizing
```{r eval=FALSE}
 h <- hist(as.matrix(assays(primed.gs)$expdata),
           breaks = 800,
           plot = FALSE)
 plot(h,
      freq = FALSE,
    xlim = c(0.3, 0.9),
    ylim = c(0, 0.5),
      main = "Histogram of gene expression in Reprogramming Study",
      xlab = "Gene expression",
      col = "darkgoldenrod2",
      border = "grey") +
 abline(v = 0.46, col = "blue")
```

### Binarize
```{r eval=FALSE}
reference.gs    <- binarize_exp(reference.gs, fix_cutoff = TRUE, binarize_cutoff = 0.65)
fibroblast.gs   <- binarize_exp(fibroblast.gs, fix_cutoff = TRUE, binarize_cutoff = 0.65)
mixed.gs        <- binarize_exp(mixed.gs, fix_cutoff = TRUE, binarize_cutoff = 0.75)
early_primed.gs <- binarize_exp(early_primed.gs, fix_cutoff = TRUE, binarize_cutoff = 0.49)
primed.gs       <- binarize_exp(primed.gs, fix_cutoff = TRUE, binarize_cutoff = 0.4555)
```

### CHECKPOINT
##### **with how long binerizing takes you may want to choose to save objects for future use.
```{r eval=FALSE}
 saveRDS(reference.gs   , "~/R Packages/mini_data/Binerized/reference.rds")
 saveRDS(fibroblast.gs  , "~/R Packages/mini_data/Binerized/fibroblast.rds")
 saveRDS(mixed.gs       , "~/R Packages/mini_data/Binerized/mixed.rds")
 saveRDS(early_primed.gs, "~/R Packages/mini_data/Binerized/early_primed.rds")
 saveRDS(primed.gs      , "~/R Packages/mini_data/Binerized/primed.rds")

 reference.gs    <- readRDS("~/R Packages/mini_data/Binerized/reference.rds")
 fibroblast.gs   <- readRDS("~/R Packages/mini_data/Binerized/fibroblast.rds")
 mixed.gs        <- readRDS("~/R Packages/mini_data/Binerized/mixed.rds")
 early_primed.gs <- readRDS("~/R Packages/mini_data/Binerized/early_primed.rds")
 primed.gs       <- readRDS("~/R Packages/mini_data/Binerized/primed.rds")
```

### fit logistic regression and find the switching pseudo-time point for each gene
```{r eval=FALSE}
reference.gs <- find_switch_logistic_fastglm(reference.gs, downsample = FALSE, show_warning = FALSE)
```

#### **This is another time consuming process, may want to choose to save objects for future use.**
```{r eval=FALSE}
saveRDS(reference_glm.gs   , "~/R Packages/mini_data/Binerized/reference_glm.rds")
reference.gs    <- readRDS("~/R Packages/mini_data/Binerized/reference_glm.rds")
```

### Filter to only include Switching Genes
```{r eval=FALSE}
reference.sg <- filter_switchgenes(reference.gs, allgenes = TRUE,r2cutoff = 0.03)
```

##### View all of the switching genes
```{r eval=FALSE}
plot_timeline_ggplot(reference.sg, timedata = colData(reference_glm.gs)$Pseudotime, txtsize = 3)
```

# Using GeneSwitchesScorer

### Filter for an even distribution
```{r eval=FALSE}
gss.sg <- select_evenly_distributed_switching_genes(reference.sg, min_time_spacing = 5)
```

##### View the selected switching genes
```{r eval=FALSE}
plot_timeline_ggplot(gss.sg, timedata = colData(reference_glm.gs)$Pseudotime, txtsize = 3)
```

### Reduce the binary counts matricies of the query data to only include the selection of evenly distributed genes from the refernence.
```{r eval=FALSE}
fibroblast_reduced      <- filter_gene_expression_for_switching_genes(fibroblast.gs@assays@data@listData$binary   , gss.sg)
mixed_reduced           <- filter_gene_expression_for_switching_genes(mixed.gs@assays@data@listData$binary        , gss.sg)
early_primed_reduced    <- filter_gene_expression_for_switching_genes(early_primed.gs@assays@data@listData$binary , gss.sg)
primed_reduced          <- filter_gene_expression_for_switching_genes(primed.gs@assays@data@listData$binary       , gss.sg)
reference_reduced       <- filter_gene_expression_for_switching_genes(reference.gs@assays@data@listData$binary    , gss.sg)
```

### Produce an estimate for the position on trajectory of each gene in each cell of a sample.
```{r eval=FALSE}
fibroblast.gss   <- create_racing_lines(fibroblast_reduced  , gss.sg)
mixed.gss        <- create_racing_lines(mixed_reduced       , gss.sg)
early_primed.gss <- create_racing_lines(early_primed_reduced, gss.sg)
primed.gss       <- create_racing_lines(primed_reduced      , gss.sg)
reference.gss    <- create_racing_lines(reference_reduced   , gss.sg)
```

### Accuracy
#### *As our samples are subsets of the reference dataset we can calculate the accuracy of GSS*
```{r eval=FALSE}
fibroblast_accuracy   <- score_gss_accuracy(fibroblast.gss, reference.sce)
mixed_accuracy        <- score_gss_accuracy(mixed.gss, reference.sce)
early_primed_accuracy <- score_gss_accuracy(early_primed.gss, reference.sce)
primed_accuracy       <- score_gss_accuracy(primed.gss, reference.sce)
reference_accuracy    <- score_gss_accuracy(reference.gss, reference.sce)
```

## Plotting

### Plot the predicted position of each sample:
#### *Optional: include the switching point of some genes of interest:*
```{r eval=FALSE}
gss_output_plot(fibroblast.gss, col = "red", overlay=FALSE, label = "fibroblast", genes_of_interest = c("ENSG00000248605","ENSG00000272168"))
gss_output_plot(mixed.gss, col = "peachpuff4", overlay=TRUE, label = "mixed")
gss_output_plot(early_primed.gss, col = "purple", overlay=TRUE, label = "early primed")
gss_output_plot(primed.gss, col = "darkgreen", overlay=TRUE, label = "primed" , genes_of_interest = c("ENSG00000236924"))
```

## Further analysis of the results

### Investigate the predicted position of indervidual cells:

#### Plot the preducted position of all cells in a Sample
###### *only run this on a selection of cells or a small sample*
```{r eval=FALSE}
for (c in 1:nrow(sc.gss$cells_flat)) {
  plot(x = 1:100, y = sc.gss$cells_flat[c,], type = "h",
       xlab = "Pseudotime Index", ylab = "Cell Position Likelihood",
       main = paste("Trajectory Progress of", rownames(sc.gss$cells_flat)[c]))
  segments(which.max(sc.gss$cells_flat[c,]),
           0.0,
           which.max(sc.gss$cells_flat[c,]),
           as.numeric(sc.gss$cells_flat[c,which.max(sc.gss$cells_flat[c,])]),
           lwd=2,
           col ="red")

}
```

#### Plot the "racing lines" for indervidual cells.
```{r eval=FALSE}
racinglines_timeline(reference.sg, reference_reduced, cell_idx = 1)
```

### Plot the distribution of inaccuracy of each sample:
```{r eval=FALSE}
hist(fibroblast_accuracy$inaccuracy  , breaks = 100)
hist(mixed_accuracy$inaccuracy       , breaks = 100)
hist(early_primed_accuracy$inaccuracy, breaks = 100)
hist(primed_accuracy$inaccuracy      , breaks = 100)
hist(reference_accuracy$inaccuracy   , breaks = 100)
```

### Investigate the predicted position of indervidual cells:
```{r eval=FALSE}
#
plot(fibroblast_accuracy$true_position_of_cells_timeIDX, fibroblast_accuracy$inaccuracy,
     xlab = "True Position", ylab = "Inaccuracy", main = "Inaccuracy by True Positions")

#
boxplot(fibroblast_accuracy$inaccuracy ~ fibroblast_accuracy$true_position_of_cells_timeIDX,
        xlab = "True Position", ylab = "Inaccuracy", main = "Boxplots of Inaccuracy by True Positions")
true_position_bin <- cut(fibroblast_accuracy$true_position_of_cells_timeIDX,
                             breaks = c(0,5,10,15,20,25,30,35,40,45,50,55,60,65,70,75,80.85,90,95,100))
boxplot(fibroblast_accuracy$inaccuracy ~ true_position_bin,
        xlab = "True Position", ylab = "Inaccuracy", main = "Boxplots of Inaccuracy by True Positions")

#
plot(fibroblast_accuracy$true_position_of_cells_timeIDX,fibroblast_accuracy$predicted_position_of_cells_timeIDX,
     xlab = "True Position", ylab = "Predicted Position", main = "Predicted Positions by True Positions")
segments(0,0,100,100, lwd=2, col ="green")
segments(0,10,100,110, lwd=2, col ="blue")
segments(0,-10,100,90, lwd=2, col ="blue")
segments(0,40,100,140, lwd=2, col ="red")
segments(0,-40,100,60, lwd=2, col ="red")

#
boxplot(fibroblast_accuracy$predicted_position_of_cells_timeIDX ~ fibroblast_accuracy$true_position_of_cells_timeIDX,
        xlab = "True Position", ylab = "Predicted Position", main = "Boxplots of Predicted Positions by True Positions")
segments(0,0,100,100, lwd=2, col ="green")
segments(0,10,100,110, lwd=2, col ="blue")
segments(0,-10,100,90, lwd=2, col ="blue")
segments(0,40,100,140, lwd=2, col ="red")
segments(0,-40,100,60, lwd=2, col ="red")
```

### Investigate the precision of min_time_spacing values:
```{r eval=FALSE}
precision <- data.frame(
  min_time_spacing = 0:100,
  n_genes          = NA,                                                   # Number of genes after min_time_spacing
  accuracy_min     = NA,                                                   # Placeholder for accuracy minimum values
  accuracy_1       = NA,
  accuracy_median  = NA,
  accuracy_mean    = NA,
  accuracy_3       = NA,
  accuracy_max     = NA
  )



for (i in 0:100){
# Filter for an even distribution
gss_genes <- select_evenly_distributed_switching_genes(reference.gs, min_time_spacing = i)

# Reduce the binary counts matricies of the query data to only include the selection of evenly distributed genes from the refernence.
sample_reduced      <- filter_gene_expression_for_switching_genes(sample.gs@assays@data@listData$binary   , gss_genes)

#
sample.gss <- create_racing_lines(sample_reduced, gss_genes)

#
accuracy <- score_gss_accuracy(sample.gss, reference.sce)

#
precision$n_genes[i+1]          <- dim(gss_genes)[1]
precision$accuracy_min[i+1]     <- summary(accuracy$accuracy)[1]
precision$accuracy_1[i+1]       <- summary(accuracy$accuracy)[2]
precision$accuracy_median[i+1]  <- summary(accuracy$accuracy)[3]
precision$accuracy_mean[i+1]    <- summary(accuracy$accuracy)[4]
precision$accuracy_3[i+1]       <- summary(accuracy$accuracy)[5]
precision$accuracy_max[i+1]     <- summary(accuracy$accuracy)[6]

#
cat(i," done \n")
}


ggplot(data = precision, mapping = aes(x = min_time_spacing)) +
  geom_point(aes(y = accuracy_min), color = "lightblue") +
  geom_point(aes(y = accuracy_1), color = "lightpink") +
  geom_point(aes(y = accuracy_median), color = "red") +
  geom_point(aes(y = accuracy_mean), color = "blue") +
  geom_point(aes(y = accuracy_3), color = "darkgreen") +
  geom_point(aes(y = accuracy_max), color = "lightblue") +
  ylab("innacuracy")

ggplot(data = precision, mapping = aes(x = n_genes)) +
  geom_point(aes(y = accuracy_min), color = "lightblue") +
  geom_point(aes(y = accuracy_1), color = "lightpink") +
  geom_point(aes(y = accuracy_median), color = "red") +
  geom_point(aes(y = accuracy_mean), color = "blue") +
  geom_point(aes(y = accuracy_3), color = "darkgreen") +
  geom_point(aes(y = accuracy_max), color = "lightblue") +
  ylab("innacuracy") + scale_x_log10()


ggplot(data = precision, mapping = aes(x = min_time_spacing)) +
  geom_point(aes(y = n_genes), color = "red") +
  scale_y_log10()
```
